---
title: "Multinomial Regression"
author: "Chris Ryan, Jeb Brown, Varsha Manickam, Meghna Kommoju, Zhennan Shen"
date: "2025-04-15"
fontsize: 9pt
classoption: "aspectratio=169"
output: 
  beamer_presentation: 
    fonttheme: professionalfonts
    highlight: kate
    theme: Rochester
header-includes:
- \definecolor{VTmaroon}{HTML}{861F41}
- \usecolortheme[named=VTmaroon]{structure}
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE,
                      comment = NA)

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})


```

## Multinomial Regression

Multinomial Regression (also called Multinomial Logistic Regression) is a classification method that is used to predict nominal outcome variables that have more than two outcomes (polytomous) that do not have a rank or order.
Examples of this would be:

- Classifying flower species (setosa, virginica, veritosa)
- Classifying student programs (academic, vocation, general)
- Classifying student majors (engineering, business, science)

## Assumptions for Multinomial Regression

- Independence of observations
- Categories of the outcome variable must be mutually exclusive and exhaustive
- No multicollinearity between independent variables
- Linear relationship between continuous variables and the logit transformation of the outcome variable

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

Do you see the above `##`?  I hate those, you need to figure out how to remove those.


## Slide with Plot


```{r pressure}
plot(pressure)
```


